---
title: "Class15"
author: 'Jaquish (PID: A59010386)'
date: "11/17/2021"
output: github_document
  
---

# Background

Today we examine a published RNA-seq experiment where airway smooth muscle cells were treated with dexamethasane, a synthetic glucocorticoid steroid with anti-inflammatory effects. (Himes et al. 2014)

## Load the contData and colData

We need 2 things
- 1: count Data
- 2: colData (the metadata that tells us about the design of the experiment)

```{r}
counts <- read.csv("airway_scaledcounts.csv", row.names=1)
metadata <-  read.csv("airway_metadata.csv")
```

```{r}
#head(counts)
```

```{r}
head(metadata)
```


Side note: 
Let's check the correspondence of the metadata and count data setup.

```{r}
metadata$id
```

```{r}
colnames(counts)
```

We can use '==' thing to see if they are the same

```{r}
metadata$id == colnames(counts)
```

Checks to see if everything is what you thought, in this case confirming there are no falses.

```{r}
all( c(T,T,T,F))
```

## Compare control to treated

First we need to access all the controls columns in our counts data.

```{r}
control.inds <- metadata$dex == "control"
control.ids <- metadata[ control.inds,]$id
```

Use these ids to access just the control columns of our 'counts' data

```{r}
head(counts[ , control.ids])
```

```{r}
control.mean <- rowMeans(counts[ , control.ids])
head(control.mean)
```

Do the same for the treated.

```{r}
treated.inds <- metadata$dex == "treated"
treated.ids <- metadata[ treated.inds,]$id
```

```{r}
head(counts[ , treated.ids])
```

```{r}
treated.mean <- rowMeans(counts[ , treated.ids])
head(treated.mean)
```

```{r}
meancounts <- data.frame(control.mean, treated.mean)
```

There are `r nrow(counts)` rows/genes in this dataset.

```{r}
nrow(counts)
```

>Q1. 38,694 genes

## Compare the control v. treated

A quick plot of our progress so far.

```{r}
plot(meancounts)
```

Each dot is a gene. Dots along the linear diagnol see no expression change. Below the diagnol is negative, above positive/increased in treated.

This would benefit from a log transform! Let's plot on a log sclae

```{r}
plot(meancounts, log="xy")
```

We often use log transformations to make data more comprehensible/sensible.

```{r}
log2(40/20)
```

Fold change of 1 if the treatment doubled in expression.

```{r}
log2(10/20)
```

Fold change of -1 if the treatment expression decreased to half of control.

Cool, lets calculate the fold change of our control v. treated.

```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```

We need to drop the zero count genes/rows..

```{r}
head(meancounts[,1:2] == 0)
```

The `which()` function tells us the indices of TRUE entries in a logical vector.

```{r}
which( c(T,F,T))
```

For our case we don't want to know the True positions within the vector, we want to know which rows have True. 

```{r}
inds <- which(meancounts[,1:2] == 0, arr.ind=TRUE)
head(inds)

```

I only care about the rows here (if there is a 0 in any column I will exclude this eventually)

```{r}
to.rm <-  unique(sort(inds[, "row"]))
head(meancounts[-to.rm])
mycounts <- meancounts[-to.rm,]
```

We now have `r nrow(mycounts)` genes remaining.

```{r}
nrow(mycounts)
```

Fold-change threshold of +2 or greater?

```{r}
sum(mycounts$log2fc > +2)
```

What percentage is this?

```{r}
round((sum(mycounts$log2fc > +2) / nrow(mycounts)) *100, 2)
```

```{r}
sum(mycounts < -2)
```

```{r}
library(DESeq2)
```

We first need to setup the DESeq input object.

```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex)
```

Run the DESeq analysis pipeline.


```{r}
dds <- DESeq(dds)
```

```{r}
res <- results(dds)
head(res)
```

Focus on the genes with a good p-value(low): plot the p-value against the log2 with volcano plot

# A Volcano Plot

This is a very common data viz of this type of data that does not really look like a volcano.

```{r}
plot(res$log2FoldChange, -log(res$padj))
abline(v=c(-2,2), col="gray")
abline(h=-log(0.05), col="gray")
```


## Adding annotation data

We want to add meaningful gene names to our dataset to determine which are up/down most significantly. 

For this we will use 2 bioconductor packages, one does the work and is called AnnotationDbi the other contains the data we are going to map between and is called "org.Hs.eg.db"  

`BiocManager::install("org.Hs.eg.db")`
`BiocManager::install("AnnotationDbi)`


```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```

```{r}
columns(org.Hs.eg.db)
```

Here we map to "SYMBOL" the common gene name that the world understands and wants.


```{r}
res$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(res), 
                     keytype="ENSEMBL",        
                     column="SYMBOL",          
                     multiVals="first")

```

```{r}
head(res)
```


# Lets finally save our results to date.

```{r}
write.csv(res, file= "allmyresults.csv")
```


# Pathway Analysis

Let's try to bring some biology insight back into this.
We will start wtih KEGG.

```{r}
library(pathview)
library(gage)
library(gageData)

data(kegg.sets.hs)

# Examine the first 2 pathways in this kegg set for humans
head(kegg.sets.hs, 2)
```

The above genes are in Entrez format. 

Before we can use KEGG we need to get our gene identifiers in the correct format for KEGG, which is ENTREZ format in this case.

```{r}
head(rownames(res))
```

```{r}
columns(org.Hs.eg.db)
```



```{r}
res$entrez <- mapIds(org.Hs.eg.db,
                     keys=row.names(res), 
                     keytype="ENSEMBL",        
                     column="ENTREZID",          
                     multiVals="first")

res$genename <- mapIds(org.Hs.eg.db,
                     keys=row.names(res), 
                     keytype="ENSEMBL",        
                     column="GENENAME",          
                     multiVals="first")
```

The main `gage()` function requires a named vector of fold changes, where the names of the values are the Entrez gene IDs.

Note that we used the mapIDs() function above to obtain Entrez gene IDs (stored in res$entrez) and we have the fold change results from DESeq2 analysis (stored in res$log2FoldChange).

```{r}
foldchanges = res$log2FoldChange
```



```{r}

names(foldchanges) = res$entrez
head(foldchanges)
```

```{r}
keggres = gage(foldchanges, gsets=kegg.sets.hs)
```

```{r}
attributes(keggres)
```

```{r}
head(keggres$less, 3)
```

The pathway() function will add out genes to a KEGG pathway as colored entries:

```{r}
pathview(gene.data=foldchanges, pathway.id="hsa05310")
```

1. Data import/read
countdata
coldata(metadata)

2. PCA (QC) 
if happy then...

3. Deseq analysis (DESeq() function)

4. Figures: Volcano Plot (ggplot or plot function)

5. Annotation (use biology to understand the up/down regulated)
KEGG, GO, etc.

6. Pathway Analysis (gage() function)









