---
title: "class09_mini_project"
author: 'Jaquish (PID: A59010386)'
date: "10/27/2021"
output:
  pdf_document: default
  html_document: default
---

```{r}
fna.data <- "WisconsinCancer.csv"
wisc.df <- read.csv(fna.data, row.names=1)
```


```{r}
# wisc.df
```
```{r}
wisc.data <- wisc.df[,-1]
```

```{r}
diagnosis <- as.factor(wisc.df$diagnosis)
#diagnosis
```

>Q1. How many observations are in this dataset?

```{r}
nrow(wisc.data)
```



>Q2. How many of the observations have a malignant diagnosis?

```{r}
sum(diagnosis == "M")
```

```{r}
table(diagnosis)
```


>Q3. How many variables/features in the data are suffixed with _mean?

```{r}
grep("_mean", colnames(wisc.data), value=TRUE)
```


```{r}
length(grep("_mean", colnames(wisc.data), TRUE))
```

```{r}
#colMeans(wisc.data)
```

```{r}
#apply(wisc.data,2,sd)
```

```{r}
wisc.pr <- prcomp( wisc.data, scale=TRUE)
```

```{r}
summary(wisc.pr)
```

>Q4. From your results, what proportion of the original variance is captured by the first principal components (PC1)?

The first PC1 captures 44% of the original variance.

>Q5. How many principal components (PCs) are required to describe at least 70% of the original variance in the data?

To get 70% you need to go up to PC3.

>Q6. How many principal components (PCs) are required to describe at least 90% of the original variance in the data?

To hit 90% you need PC7.

>Q7. What stands out to you about this plot? Is it easy or difficult to understand? Why?

Very messy and difficult to understand because there is just too much data.

```{r}
#biplot(wisc.pr)
```

>Q8. Generate a similar plot for principal components 1 and 3. What do you notice about these plots?

They are much prettier and look pretty similar. The plot with 1 and 3 look to be tighter together because they cover a small variance.  

```{r}
plot(wisc.pr$x[,1:2], col= diagnosis, xlab = "PC1", ylab = "PC2")
```

```{r}
plot(wisc.pr$x[,1], wisc.pr$x[,3], col= diagnosis, xlab = "PC1", ylab = "PC3")
```

```{r}
df <- as.data.frame(wisc.pr$x)
df$diagnosis <- diagnosis
library(ggplot2)
```

```{r}
ggplot(df) + 
  aes(PC1, PC2, col= diagnosis) + 
  geom_point()
```

```{r}
pr.var <- wisc.pr$sdev^2
pr.var
?head(pr.var)
```

```{r}
pve <-  pr.var/ sum(pr.var)
pve
```

```{r}
plot(pve, xlab = "Principal Component", 
     ylab = "Proportion of Variance Explained", 
     ylim = c(0, 1), type = "o")
```

```{r}
barplot(pve, ylab = "Precent of Variance Explained",
     names.arg=paste0("PC",1:length(pve)), las=2, axes = FALSE)
axis(2, at=pve, labels=round(pve,2)*100 )
```

Optional:
```{r}
#install.packages("factoextra")
#library(factoextra)
#fviz_eig(wisc.pr, addlabels = TRUE)
```


>Q9. For the first principal component, what is the component of the loading vector (i.e. wisc.pr$rotation[,1]) for the feature concave.points_mean?

The concave.points_mean is -0.26. This value is the average of the downward curve PC1 took because of the distance from this point.  

```{r}
#wisc.pr$rotation[,1]
```



>Q10. What is the minimum number of principal components required to explain 80% of the variance of the data?

You need 4 PC's to describe 80% of the variance.

```{r}
var <- summary(wisc.pr)
sum(var$importance[3,] <0.8)
```

```{r}
data.scaled <- scale(wisc.data)
```

```{r}
data.dist <- dist(data.scaled)
```

```{r}
#data.dist
```

```{r}
wisc.hclust <- hclust(data.dist, "complete")
```

```{r}
plot(wisc.hclust)
abline( h=15, col="red", lty=2)
```


```{r}
wisc.hclust.clusters <- cutree(wisc.hclust,k=5)
table(wisc.hclust.clusters, diagnosis)
```

>Q12. Can you find a better cluster vs diagnoses match by cutting into a different number of clusters between 2 and 10?

Cluster 5 seems to be good because there is a nice separation between the diagnoses of benign and malignant. 

>Q13. Which method gives your favorite results for the same data.dist dataset? Explain your reasoning.

I also like the ward.D2 because the groups look very clean and it builds it so the variance is minimized. This is similar to why I chose 5 clusters above because it groups them together into a single category best as possible. 

```{r}
wisc.hclust <- hclust(data.dist, "ward.D2")
plot(wisc.hclust)
```

