---
title: "Halloween project"
author: 'Jaquish (PID: A59010386)'
date: "10/29/2021"
output:
  pdf_document:
    latex_engine: xelatex
---

```{r}
#read.csv("candy-data.csv")
candy_file <- "candy-data.csv"
```

```{r}
candy = read.csv("candy-data.csv", row.names=1)
head(candy)
```


```{r}
row.names(candy)
```

```{r}
nrow(candy)
```


>Q1. How many different candy types are in this dataset?

85 different types of candy.

> Q2. How many fruity candy types are in the dataset?

38 fruity candies

```{r}
colnames(candy)
```


```{r}
sum(candy$fruity)
```


> Q3. What is your favorite candy in the dataset and what is it’s winpercent value?

My favorite is Butterfingers but they are not on the list so I"ll go with Nerds at only 55%.
```{r}
candy["Nerds", ]$winpercent
```


>Q4. What is the winpercent value for “Kit Kat”?

76%

```{r}
candy["Kit Kat", ]$winpercent
```


>Q5. What is the winpercent value for “Tootsie Roll Snack Bars”?

```{r}
candy["Tootsie Roll Snack Bars", ]$winpercent
```


```{r}
library("skimr")


```

```{r}
skim(candy)
```

> Q6. Is there any variable/column that looks to be on a different scale to the majority of the other columns in the dataset?

The winpercent is on a different scale.

> Q7. What do you think a zero and one represent for the candy$chocolate column?

A 1 means that it was chosen over another candy while the zero means it lost against another candy.

```{r}
hist(candy$winpercent)
```

> Q9. Is the distribution of winpercent values symmetrical?

No not symmetrical.

> Q10. Is the center of the distribution above or below 50%?

The center is below 50%.

> Q11. On average is chocolate candy higher or lower ranked than fruit candy?

The average of all winpercent values for chocolate is ranked higher than that of fruity.

```{r}
chocolate <- candy[ as.logical(candy$chocolate), ]$winpercent
mean(chocolate)
```

```{r}
fruity <- candy[ as.logical(candy$fruity), ]$winpercent
mean(fruity)
```

> Q12. Is this difference statistically significant?

```{r}
t.test(chocolate, fruity)
```

> Q12. Is this difference statistically significant?

It is signigicant based on p-value.


>Q13. What are the five least liked candy types in this set?

```{r}
head(candy[order(candy$winpercent),], n=5)
```


```{r}
rownames(head(candy[order(candy$winpercent,decreasing=TRUE),], n=5))
```


```{r}
#library(dplyr)

#candy %>% arrange(desc(winpercent)) %>% head(5)

head(candy[order(candy$winpercent),],decreasing=TRUE, n=5)
```


```{r}
my_cols=rep("black", nrow(candy))
my_cols[as.logical(candy$chocolate)] = "chocolate"
my_cols[as.logical(candy$bar)] = "brown"
my_cols[as.logical(candy$fruity)] = "pink"
```


```{r}
library(ggplot2)

ggplot(candy) + 
  aes(winpercent, reorder(rownames(candy),winpercent)) +
  geom_col(fill=my_cols)
```
> Q17. What is the worst ranked chocolate candy?

Nik L Lip

> Q18. What is the best ranked fruity candy?

Starburst


```{r}
library(ggrepel)
```

# How about a plot of price vs win

```{r}
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols, size=3.3, max.overlaps = 5)
```

# change my fruity color to red

```{r}
my_cols[as.logical(candy$fruity)] <- "red"
ggplot(candy) +
  aes(winpercent, pricepercent, label=rownames(candy)) +
  geom_point(col=my_cols) + 
  geom_text_repel(col=my_cols)
```

```{r}
gsub("Ö", "'", rownames(candy))
```


> Q19. Which candy type is the highest ranked in terms of winpercent for the least money - i.e. offers the most bang for your buck?

From this list it looks like Reeses Miniatures is the best bang for buck with high popularity.

```{r}
ord <- order(candy$winpercent,decreasing=TRUE)
head( candy[ord,c(11,12)], n=5 )
```



> Q20. What are the top 5 most expensive candy types in the dataset and of these which is the least popular?

Nik L Nips are most expensive and least popular. 

```{r}
ord <- order(candy$pricepercent, decreasing = TRUE)
head( candy[ord,c(11,12)], n=5 )
```

```{r}
library(corrplot)
```

```{r}
cij <- cor(candy)
corrplot(cij)
```

> Q22. Examining this plot what two variables are anti-correlated (i.e. have minus values)?

Fruity and chocolate are not correlated. Bar and pluribus as well.

> Q23. Similarly, what two variables are most positively correlated?

Chocolate and bar. Chocolate and winpercent. Chocolate and pricepercent.

```{r}
pca <- prcomp(candy, scale=TRUE)
summary(pca)
```

```{r}
plot(pca$x[,1:2])
```

```{r}
plot(pca$x[,1:2], col=my_cols, pch=16)
```

```{r}
# Make a new data-frame with our PCA results and candy data
my_data <- cbind(candy, pca$x[,1:3])
```

```{r}
p <- ggplot(my_data) + 
        aes(x=PC1, y=PC2, 
            size=winpercent/100,  
            text=rownames(my_data),
            label=rownames(my_data)) +
        geom_point(col=my_cols)

p
```

```{r}
#library(ggrepel)

p + geom_text_repel(size=3.3, col=my_cols, max.overlaps = 7)  + 
  theme(legend.position = "none") +
  labs(title="Halloween Candy PCA Space",
       subtitle="Colored by type: chocolate bar (dark brown), chocolate other (light brown), fruity (red), other (black)",
       caption="Data from 538")
```

```{r}
par(mar=c(8,4,2,2))
barplot(pca$rotation[,1], las=2, ylab="PC1 Contribution")
```

> Q24. What original variables are picked up strongly by PC1 in the positive direction? Do these make sense to you?

The fruity, hard and pluribus. This makes sense because most fruity candy tends to come in bags and are hard.